Typed Functional Genetic Programming

### 2013-05-08 tex/thesis.tex
[tex] proofs of barendrecht inference rules ...
[tex] tree representations of terms and solving of Inhab tree

\section{Lambda term}

@@

	\begin{align*}
		 x   \in V                 &\Rightarrow     x   \in \Lambda \\
		 c   \in C                 &\Rightarrow     c   \in \Lambda \\
		 M,N \in \Lambda           &\Rightarrow ( M N ) \in \Lambda \\
		 x   \in V , M \in \Lambda &\Rightarrow ( \lambda x . M ) \in \Lambda 
	\end{align*}

	\textbf{TODO} 
	\begin{itemize} 
		\item TALK ABOUT "parenthesis" conventions. 
	\end{itemize}

@@

Inference rule 1: 
\[
	\frac{\Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} \vdash M : \alpha }
	     {\Gamma \vdash (\lambda x_1 \dots x_n . M) : 
	     \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha }
\]
\\
... there is a grammar rule:
\[ 
	( \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha , \Gamma )  \longmapsto
	\bigg( \mbox{ {\Large 
	$\lambda x_1 \dots x_n .$ 
	}}( \alpha , \Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} ) \quad \bigg)
\]
\\

Inference rule 2: 
\[
	\frac{ (f , \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha ) \in \Gamma \qquad
	       \Gamma \vdash M_1 : \rho_1 \quad
	       \dotsm \quad
	       \Gamma \vdash M_m : \rho_m        
	      }
	     {\Gamma \vdash (f M_1 \dots M_m) : \alpha}
\]
\\
... there is a grammar rule:
\[ 
	( \alpha , \Gamma )  \longmapsto
	\bigg( \mbox{ {\Large f }}
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_1 , \Gamma )
	  \mbox{ \Vtextvisiblespace[1em] } 
	  \dots
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_m , \Gamma )
	  \quad \bigg)
\]
\\\\
\textbf{TODO} 
\begin{itemize}
	\item SHOW correctness of those inference rules by composing them of 
		  $E^{\rightarrow}$, $I^{\rightarrow}$ and \textit{axiom}.
	\item SHOW more examples of inference rules transformed into grammar rules.
	\item DESCRIBE general algorithm for this transformation.
	\item TALK ABOUT $\tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha$ 
	\item TALK ABOUT $\beta \eta$-normal form which is generated by this method.
\end{itemize}

@@

~\\\\
\textbf{TODO}: TALK ABOUT : By \textit{solving} And-or tree for Inhabitation tree
               we obtain that typed tree notation for typed $\lambda$-term,
               which must be firstly INTRODUCED.   \\
\textbf{TODO}: TALK (more?) ABOUT "Barendregt-like" subsection \ref{barlike}
\\\\

@@

\subsection{Inhabitation Machine}

>>>>

\section{Lambda term}
\label{deflam}

@@

	\begin{align*}
		 x   \in V \cup C          &\Rightarrow     x   \in \Lambda \\
		 M,N \in \Lambda           &\Rightarrow ( M N ) \in \Lambda \\
		 x   \in V , M \in \Lambda &\Rightarrow ( \lambda x . M ) \in \Lambda 
	\end{align*}

	\textbf{TODO} 
	\begin{itemize} 
		\item TALK ABOUT "parenthesis" conventions (and packing of lambda abstractions). 
	\end{itemize}

@@

Inference rule 1: 
\[
	\frac{\Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} \vdash M : \alpha }
	     {\Gamma \vdash (\lambda x_1 \dots x_n . M) : 
	     \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha }
\]
\\
Proof of correctness:
\[
	\dfrac{
		\dfrac
		 {\Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} \vdash M : \alpha}
		 {\dfrac
		   {\Gamma \cup \{ (x_1,\tau_1),\dots,(x_{n-1},\tau_{n-1})\} \cup 
		                \{(x_n,\tau_n) \} \vdash M : \alpha}
		   {\dfrac{\Gamma \cup \{ (x_1,\tau_1),\dots,(x_{n-1},\tau_{n-1})\}  
		                \vdash (\lambda x_n . M) : \tau_n \rightarrow \alpha}
				  { \vdots }		   
		   }
		 }		 
	 }
	     {\Gamma \vdash (\lambda x_1 \dots x_n . M) : 
	     \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha }
\]
\\
... there is a grammar rule:
\[ 
	( \tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha , \Gamma )  \longmapsto
	\bigg( \mbox{ {\Large 
	$\lambda x_1 \dots x_n .$ 
	}}( \alpha , \Gamma \cup \{ (x_1,\tau_1),\dots,(x_n,\tau_n) \} ) \quad \bigg)
\]
\\

Inference rule 2: 
\[
	\frac{ (f , \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha ) \in \Gamma \qquad
	       \Gamma \vdash M_1 : \rho_1 \quad
	       \dotsm \quad
	       \Gamma \vdash M_m : \rho_m        
	      }
	     {\Gamma \vdash (f M_1 \dots M_m) : \alpha}
\]
\\
Proof of correctness (\textbf{TODO REPAIR} Conceptually it is ok but there is sazba-bug somewhere): 
\[
   \dfrac
     {\dfrac
      {\dfrac
       {\dfrac         
         {\dfrac  
          {\dfrac
           {\boxed{(f , \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha ) \in \Gamma}}
           {\Gamma \vdash f : \rho_1 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha}
           \quad
           \boxed{\Gamma \vdash M_1 : \rho_1} }
          {\Gamma \vdash (f M_1) : \rho_2 \rightarrow \dots \rightarrow \rho_m \rightarrow \alpha }
          }{\vdots} 
         \quad 
         \ddots }
       {\Gamma \vdash (f M_1 \dots M_{m-2}) : \rho_{m-1} \rightarrow \rho_m \rightarrow \alpha}
       \quad
       \boxed{\Gamma \vdash M_{m-1} : \rho_{m-1}}  }
      {\Gamma \vdash (f M_1 \dots M_{m-1}) : \rho_m \rightarrow \alpha}       
      \quad 
      \boxed{\Gamma \vdash M_m : \rho_m} }
	 {\Gamma \vdash (f M_1 \dots M_m) : \alpha}
\]
\\
... there is a grammar rule:
\[ 
	( \alpha , \Gamma )  \longmapsto
	\bigg( \mbox{ {\Large f }}
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_1 , \Gamma )
	  \mbox{ \Vtextvisiblespace[1em] } 
	  \dots
	  \mbox{ \Vtextvisiblespace[1em] } 
	  ( \rho_m , \Gamma )
	  \quad \bigg)
\]
\\\\
\textbf{TODO} 
\begin{itemize}
	\item SHOW correctness of those inference rules by composing them of 
		  $E^{\rightarrow}$, $I^{\rightarrow}$ and \textit{axiom}.
	\item SHOW more examples of inference rules transformed into grammar rules.
	\item DESCRIBE general algorithm for this transformation.
	\item TALK ABOUT $\tau_1 \rightarrow \dots \rightarrow \tau_n \rightarrow \alpha$ 
	\item TALK ABOUT $\beta \eta$-normal form which is generated by this method.
\end{itemize}

@@

The motivation for defining \textit{solving} of a And-or tree the way we did is that
a found tree $T'$ corresponds to generated $\lambda$-term. 
In order to understand this correspondence let's now talk about various tree representations
of $\lambda$-terms.

\subsection{Tree representations of $\lambda$-terms}

From the definition of $\lambda$-term (\ref{deflam}) we can straightforwardly derive 
the classical tree representation for $\lambda$-terms. Term M is translated into tree $T[M]$ by following rules:

\begin{itemize}
	\item $x \in V \cup C$ translates into \textit{leaf} $x$.
	\item $(M N)$ translates into tree\\
		\Tree
			[.@	
		 		\text{$T[M]$}
		 		\text{$T[N]$}		 			
			] 
	\item $\lambda x . M$ translates into tree\\
	 	\Tree
			[.\text{$\lambda x$}	
		 	 	\text{$T[M]$}	
			] 
\end{itemize}

We can enhance this representation by compressing consecutive lambda abstractions into one
tree node like this:

\begin{itemize}
	\item $\lambda x_1 \dots x_n . M$ translates into tree\\
	 	\Tree
			[.\text{$\lambda x_1 \dots x_n$ }	
		 	 	\text{$T[M]$}	
			] 
\end{itemize}

As this representation comes directly from definition it is evident 
that it covers all possible $\lambda$-terms.

For representing expressions as trees it is however more common use a little different
representation. It will also be the representation suitable for showing 
that \textit{solving} Inhabitation tree generates wanted $\lambda$-term.

\begin{itemize}
    \item $x \in V \cup C$ translates into \textit{leaf} $x$.
	\item $(f M_1 M_2 \dots M_n)$ where $f \in V \cup C, n \geq 1$ translates into tree\\
		\Tree
			[.f	
		 		\text{$T[M_1]$}
		 		\text{$T[M_2]$}
		 		\text{$\dots$}
		 		\text{$T[M_n]$}		 				 			
			] 
	\item $\lambda x_1 \dots x_n . M$ translates into tree\\
	 	\Tree
			[.\text{$\lambda x_1 \dots x_n$ }	
		 	 	\text{$T[M]$}	
			] 
\end{itemize}

Notice that this representation does not cover all $\lambda$-terms, 
e.g. $(\lambda x.x) y$ is not expressible in it. But it does not bother us.  
\\\\
Let's now consider representation for \textit{typed $\lambda$-terms}.
Straightforward approach would be to add to each node a type entry which 
would be the type of the $\lambda$-term corresponding to subtree having this
node as the root node.

Approach more suitable for our purpose is to add a special type node above each node.
More specifically:

Let us consider tree $t$ corresponding to a $\lambda$-term of a type
$\sigma$ with root $r$ and subtrees $s_1 , \dots , s_n$. 
Then corresponding tree $TT[t]$ for typed $\lambda$-term is 
obtained from the tree $t$ as follows:

\begin{equation*}
\mbox{ 
TT[
\Tree
	[.r 	
	  	  \text{$s_1$}
		  \text{$s_2$}
		  \text{$\dots$}
		  \text{$s_n$}
	] 
}]=
\mbox{
\Tree
	[.\text{$\sigma$ }
	    [.r 	
	  	  \text{$TT[s_1]$}
		  \text{$TT[s_2]$}
		  \text{$\dots$}
		  \text{$TT[s_n]$}
		]	  	
	] 
}
\end{equation*}

Now we can finally put the pieces together. 
Every solution to a Inhabitation tree has this just described tree form of a typed $\lambda$-term.

~\\
\textbf{TODO} 
\begin{itemize}
\item EXAMPLES of tree representations of $\lambda$-terms  
\item TALK (more?) ABOUT "Barendregt-like" subsection \ref{barlike}\\
		Things about $\beta\eta$ normal form, etc.    
\end{itemize}

@@

\subsubsection{A* algorithm}

\subsection{Inhabitation Machine}
\textbf{TODO} 
\begin{itemize}
\item DESCRIBE Inhabitation Machine...
\end{itemize}
